---
title: "Tidyverse Presentation Exercises Solutions"
author: "Shelby Smith"
date: "10/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `ggplot2` Exercises

```{r}
library(dslabs)
head(gapminder)
```


1. To view the distribution of discrete variables like `continent`, try generating a bar chart that counts the number of observations of each `continent` in the dataset. Once you have this generated, try setting `fill = continent` within the global aesthetics & see how your visualization changes.

```{r}
ggplot(gapminder, aes(x = continent)) +
  geom_bar()
```

```{r}
ggplot(gapminder, aes(x = continent, fill = continent)) +
  geom_bar()
```

2. What is the distribution of `life_expectancy` over the whole data set? (Hint: use the `geom_density` layer, after mapping the x-variable to  `life_expectancy`)

```{r}
ggplot(gapminder, aes(x = life_expectancy)) +
  geom_density()
```

3. Can we group this by continent? Try setting "fill = `continent`" after "x = `life_expectancy`". (Hint: `geom_density(alpha = 0.5)` makes the density distributions transparent). 

```{r}
ggplot(gapminder, aes(x = life_expectancy, fill = continent)) +
  geom_density(alpha = 0.5)
```

4. How has `life_expectancy` changed over the years, `year`, by `continent`? Try fitting this first with a scatter plot, and then with a 'lm' smoother to see two different ways of visualizing this. (Hint: one is much simpler on the eye!)

```{r, message = FALSE}
ggplot(data = gapminder) +
  geom_smooth(mapping = aes(x = year, y = life_expectancy, color = continent), method = 'lm')
```

