---
title: "Tidyverse Presentation Exercises Solutions"
author: "Shelby Smith"
date: "10/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `ggplot2` Exercises

```{r}
library(dslabs)
head(gapminder)
```


1. To view the distribution of discrete variables like `continent`, try generating a bar chart that counts the number of observations of each `continent` in the dataset. Once you have this generated, try setting `fill = continent` within the global aesthetics & see how your visualization changes.

```{r}
ggplot(gapminder, aes(x = continent)) +
  geom_bar()
```

```{r}
ggplot(gapminder, aes(x = continent, fill = continent)) +
  geom_bar()
```

2. What is the distribution of `life_expectancy` over the whole data set? (Hint: use the `geom_density` layer, after mapping the x-variable to  `life_expectancy`)

```{r}
ggplot(gapminder, aes(x = life_expectancy)) +
  geom_density()
```

3. Can we group this by continent? Try setting "fill = `continent`" after "x = `life_expectancy`". (Hint: `geom_density(alpha = 0.5)` makes the density distributions transparent). 

```{r}
ggplot(gapminder, aes(x = life_expectancy, fill = continent)) +
  geom_density(alpha = 0.5)
```

4. How has `life_expectancy` changed over the years, `year`, by `continent`? Try fitting this first with a scatter plot, and then with a 'lm' smoother to see two different ways of visualizing this. (Hint: one is much simpler on the eye!)

```{r, message = FALSE}
ggplot(data = gapminder) +
  geom_smooth(mapping = aes(x = year, y = life_expectancy, color = continent), method = 'lm')
```

## Data Manipulation Exercises

1. Extract data from the *gapminder* dataset that contains the following information:
- Country: Belize
- Year: 1990
- Life_expectancy
- Population
- Continent


```{r}
gapminder %>% filter(country == c('Belize')) %>%
  filter(year == 1990) %>%
  select(country, year, life_expectancy, population, continent)

```

2. Maybe we're interested in calculating the gross domestic product per capita. Use the mutate function & write an expression to calculate this, and create a new column in the dataframe. (Hint: gdp / population)
```{r}
gapminder %>% mutate(gdp_percap = gdp / population) %>%
  select(country, year, gdp, population, gdp_percap)
```

3. Build on your code from question 2, and find out the top 3 GDP per capita in each continent
(Hint: use %>% head(3) at the end to only select the top 3). 

```{r}
# Descending gdp_percap for Africa
gapminder %>% mutate(gdp_percap = gdp / population) %>%
  filter(continent == 'Africa') %>%
  arrange(desc(gdp_percap)) %>% 
  select(continent, country, year, population, gdp, gdp_percap) %>% 
  head(3)


gapminder %>% mutate(gdp_percap = gdp / population) %>%
  filter(continent == 'Europe') %>%
  arrange(desc(gdp_percap)) %>% 
  select(continent, country, year, population, gdp, gdp_percap) %>% 
  head(3)

# To be continued for all countries!
```

### Challenge Exercises using the NYC Flights Data
  
```{r}
#install.packages('nycflights13')
library(nycflights13)
data(flights)

data(airlines)

flights_carrier <- flights %>%
  inner_join(airlines, by=c('carrier'='carrier'))

glimpse(flights_carrier)
```

1. Using the `flights_carrier` dataset, create a new data set, `flights2` that contains the carrier name, year, month, day, departure delay, arrival delay, origin airport, destination airport, and flight number.

```{r}
flights2 <- select(flights_carrier, carrier, year, month, day, dep_delay, arr_delay, origin, dest, flight)
head(flights2)
```

2. How many unique flight routes does American Airlines Inc. run departing from the New York area? (Hint: use the n_distinct() command within `summarise`)

```{r}
flights2 %>% filter(carrier == 'AA') %>% summarize(n_distinct(flight))
```


3. How many unique flight routes does American Airlines Inc. run from each of the three area airports?

```{r, message = FALSE}
flights2 %>% filter(carrier == 'AA') %>% group_by(origin) %>% summarize(n_distinct(flight))
```


4. What is the average departure delay of a American Airlines Inc. flight leaving JFK? LGA? EWR? What about leaving any New York area airport?

```{r, message = FALSE}
# Mean for each origin
flights2 %>% filter(carrier == 'UA') %>% filter(dep_delay != 'NA') %>% group_by(origin) %>% summarize(mean=mean(dep_delay))

#Mean for any New York airport
flights2 %>% filter(carrier == 'UA') %>% filter(dep_delay != 'NA') %>% summarize(mean=mean(dep_delay))
```

